Refunctionalization of Abstract Abstract Machines (Functional Pearl)

### 2018-06-30 paper/paper.tex
more revisions

%% Title information
\title{Refunctionalization of Abstract Abstract Machines (Functional Pearl)}         %% [Short Title] is optional;
                                        %% when present, will be used in
                                        %% header instead of Full Title.
%\titlenote{with title note}             %% \titlenote is optional;
                                        %% can be repeated if necessary;
                                        %% contents suppressed with 'anonymous'
\subtitle{Filling the Gap Between Abstract Abstract Machines and Abstract Definitional Interpreters}        %% \subtitle is optional
%\subtitlenote{with subtitle note}       %% \subtitlenote is optional;
                                        %% can be repeated if necessary;
                                        %% contents suppressed with 'anonymous'

@@

  In this paper, we show their functional correspondence and how to syntactically 
  transform small-step abstract abstract machines into big-step abstract definitional
  interpreters.
  The transformations include linearization, lightweight fusion, disentanglement, 
  refunctionalization, and finally transformation from CPS back to direct-style 
  with delimited control operators. 
  Linearization expresses nondeterministic choices by first-order data types, 
  after which refunctionalization transforms the first-order datatypes representing continuations
  into higher-order functions. 
  The refunctionalized AAM is an abstract interpreter written in continuation-passing style
  with two layers of continuations.

@@

Applying the same idea to big-step definitional interpreters, \citeauthor{darais2017abstracting}
built abstract definitional interpreters (ADI) that are written in monadic style
\cite{darais2017abstracting}. One of the advantages of a monadic interpreter is
that it is modular and composable. By changing the underlying monads, the definition
of the interpreter is not modified, but we can recover different semantics, including
the concrete semantics and various abstract semantics such as context-sensitivity and
abstract garbage collection \cite{Sergey:2013:MAI:2491956.2491979}.

@@

In the domain of abstract semantic artifacts, by contrast, the relations between small-step
abstract abstract machines and big-step abstract definitional interpreters, as well as the
question of deriving one from the other, are not well studied.
One of the fundamental differences between concrete semantic artifacts and abstract semantic
artifacts is that a sound instance of the latter must tell you facts about all possible execution paths, and
thus introduce nondeterminism. In addition to ensuring termination, abstract semantic
artifacts are usually equipped with a cache of reachable states; the cache is
updated in a monotonic way and guaranteed to reach the least fixed-point eventually.
In this functional pearl, we make a first step in answering the question of interderiving
abstract semantics artifacts, and provide a constructive answer to relate AAM and ADI
by presenting a series of syntactic transformations on the program from small-step
abstract abstract machines to big-step abstract definitional interpreters.

@@

  \item Section~\ref{disen} discusses \textbf{disentanglement}. In this part, we identify 
    different first-order datatypes which represent different layers of continuations, and
    disassemble their handlers into separate functions.

  \item Applying \textbf{refunctionalization} \cite{DANVY2009534, Danvy:2006:RW:2171265.2171268} 
    to disentangled AAM is shown in Section~\ref{sec:refunc}. 
    %as the left inverse of Reynolds' defunctionalization \cite{Reynold:72,Reynolds:HOSC98-revisited} 
    Refunctionalization replaces the first-order datatypes representing continuations with
    higher-order functions, and associates dispatching logic with proper higher-order functions. 
    We then derive an abstract interpreter written in continuation-passing style with two
    layers of continuations due to the nondeterminism.

@@

We can now faithfully describe the state transition function @step@,
which when given a machine state, determines its successor state.
The function @step@ is a partial function that only handles non-final states,
which must have a successor; the final case of a state is handled by funciton
@drive@, which will be mentioned in the next section.

@@

\subsection{Abstracting Abstract Machines} \label{aam}
Abstracting abstract machines (AAM) is a systematic methodology that derives sound
abstract interpreters for higher-order functional languages from concrete abstract 
machines \cite{van2012systematic, van2010abstracting}. An abstract abstract machine
implements computable abstract semantics which approximates the runtime behaviors of 
programs.
Since the state space of concrete execution is possibly infinite, the key insight of 
AAM approach when analyzing programs is to allocate both bindings and continuations 
on the store, and then bound the addresses space to be finite. Since each component
of state is finite, the abstracted machine-state space is also finite, and therefore
computable.

@@

For clarity, we divide the store into two separate stores: the binding stores @BStore@, and
the continuation store @KStore@.
The binding store maps binding addresses to sets of closure values, whereas the continuation
store maps continuation addresses to sets of continuations.
We then define a generic class @Store[K,V]@ that performs joining when updating elements
in a store (below). By parameterizing @Store[K,V]@ with @[BAddr, Storable]@ and
@[KAddr, Cont]@, we obtain @BStore@ and @KStore@, respectively.
We note that both the value store and continuation store are uodated monotonically;
it continuously grows and never shrinks.

@@

The codomain of binding stores @Storable@ is the same as previously defined for CESK 
machines. The codomain of continuation stores @Cont@, on the other hand, is comprised of
a @Frame@ object and a continuation address @KAddr@. To mimic the runtime call stack, 
@KAddr@ plays the role of representing the remaining stack frames.
But since the continuation store may contain multiple continuations, the dereferencing 
of continuation addresses is also nondeterministic.

@@

\subsubsection{Allocating Addresses}
Up to this point, we have not described allocating addresses in stores, nor handling
the time stamp @Time@. In abstract interpretation, however, these are key ingredients
to achieve analyses with different sensitivities, as well as to perform a finite
state space analysis \cite{Gilray:2016:ACP:2951913.2951936}.
To effectively approximate the runtime behavior, we introduce a finite program contours
@time@ that encodes the program execution history. The function @tick@ is used to refresh
the "time" and get a "new time". We use a list of expression (which are drawn from the
control component) to encode the calling contexts history, and as we will see in Section~
\ref{kcfainst}, by applying different @tick@ functions on the timestamp, we are able to
obtain a family of analyses.

@@

However, in contrast to the concrete CESK machine, the @drive@ function performs 
collecting semantics instead of the valuation semantics. That is, for the purpose of
analyzing programs, the function @drive@ collects all the intermediate machine states 
as the program is abstractly executing. The following code shows a variant of the 
worklist algorithm to find the fixed-point of the set of states.
Function @drive@ always applies function @step@ to the head element @hd@ of
the worklist @todo@ if @hd@ is unseen. It then inserts the result of @step@ to
the rest of worklist, and in the meantime adds @hd@ to the explored states set.
If the worklist is empty, @drive@ simply returns the set of reachable states
up to the current execution point.

@@

In this section, we describe a variant of AAM that allows the stack to be unbounded 
which uses a precise model of call stack as we did in the concrete CESK machine.
Instead of allocating continuations in the store and embedding addresses of 
continuations in states, we intend to use a list of frames to explicitly model 
the stack. By doing so, we recover the call structure as the same as runtimes
(so called pushdown control-flow analysis), but since the stack is unbounded, 
the state space is potentially infinite.

The pushdown AAM with unbounded stack is also described by \citeauthor{van2012systematic} 
\cite{van2012systematic}. 
The reason we show it here is to keep the consistency of abstract semantics during 
transformations -- since the eventual abstract definitional interpreters simply inherit
a precise control-flow from its defining language, we also would like to start from a
pushdown AAM that is also able to precisely match calls and returns.
For readers who are not familiar with pushdown analysis, we have a detailed discussion 
in Section~\ref{pdcfarevisit}.

@@

In the previous section, we show that by keeping an unabstracted stack in the state
space, we can recover the precise call/return match. Now we begin describing the
transformations step by step. Our base machine for now is the AAM with unbounded
stack, though as we will later transform the stack structure to higher-order functions
representing continuations, it does not matter what kind of AAM we start from.
This is because it would not be possible to construct a new frame on the stack, nor
to allocate continuations in the store.
Thus, our choice to start from an AAM with unbounded stack is motivated simply because
it has an equivalent stack model to abstract definitional interpreters (our final target).

The underlying semantics of AAM is fundamentally nondeterministic: there are possibly
multiple target closures when dereferencing an address, and we have to explore both branches
if the conditionals exist in the language. But the reduction context (i.e., the @Frame@ in
our program) is \emph{not} nondeterministic, and the worklist @todo@ actually implicitly
handles the nondeterminism. Thus, if we simply refunctionalize the frames to higher-order
functions, it does not help us to move toward abstract definitional interpreters.

@@

We apply this idea to AAM and our first step of transformations is to linearize all
nondeterministic choices. This step removes all nondeterminism from the @step@ function.
Likewise, the @Frame@ saves the information of the caller in concrete executions.
We then introduce another layer of control, and define a case class @NDCont@
that saves the information at a fork point when we have multiple target closures.
We also add a new field @ndk@ to the definition of state, which we now call @NDState@.
For clarity of presentation in differentiating between the two continuations, we elect to call
the first a \emph{continuation}, and the second a \emph{meta-continuation}
\footnote{Traditionally, they are also called \emph{success continuation} and \emph{failure
continuation} \cite{Danvy:1990:AC:91556.91622}.}.

@@

Here we also use first-order data types to explicitly represent the meta-continuation which
controls the nondeterminism.
Each @NDCont@ object contains a list of closures that are possible functions to be invoked,
a set of values that will be bound to the function's formal argument, and the store, time,
and frames at the fork point. In the definition of @NDState@, an auxiliary function @toState@
is added that converts itself to @State@.

@@

By atomically evaluating @f@, we obtain a set of closures, with the transition being 
deterministic in regards to the first element of the closure set.
As such, we prepare a new environment, a new store, and a new frame list only for the
first closure in that set. A new meta-continuation @new_ndk@ is also constructed,
which contains the rest of closures, the values of argument @ae@, the store, the time,
and the new frame list. We use the store before updating, because for different closures
they may form different binding addresses @baddr@. We also use the new frames, because
all closures at this fork point share the same stack and return point.

@@

If the continuation @konts@ is an empty list, then it means we have reached
the end of the computation (i.e., a halt); otherwise, we should return the values of 
@ae@ to its caller which is contained in the top frame of stack.

However, since we have added nondeterminism into state, an empty list of frames means
we have reached the halt of \textit{one computation path}, and we should determine the
next state indicated by @NDCont@. Thus, we have a pattern matching on @ndk@:

\begin{itemize}
  \item If the closure set is an empty set,
then we have tried all the closures of this fork point and should move to the
next fork point. Since the @NDCont@ is represented by a list, and we always
append new elements to its front, we are resuming to the most recent fork point by
popping up the front-most element.
  \item Otherwise, we pop a closure from the set, then build a new
environment and a new store for the body expression of the closure;
we use the same frame list and time that are copied from the fork point rather
than current one.
The meta-continuation @new_ndk@ is also updated by the removal of that closure.
\end{itemize}

@@

The @drive@ function is also changed: there is no worklist anymore, as all the
potentially unexplored states are embedded into the continuation for
nondeterminism.
The termination of the analysis occurs when we reach an @NDState@ in which the expression
is atomic and both the continuation and meta-continuation are empty lists. This corresponds 
to the case that @todo@ list is empty.

At this point, we have obtained a linearized abstract abstract machine. If we imagine that the
classical AAM explores a graph of reachable states, then the linearized AAM flattens the
graph to a linear sequence.

@@

With fusing complete, our AAM appears similar to a ``big-step'' interpreter, although it
still has the machine state representation inside. In the disentanglement transformation,
we identify the first-order data types which represent reduction contexts
and lift the code blocks that handle these reduction contexts to be individual functions.

Since there are two layers of continuations, we obtain three mutually recursive functions:
@drive_step@, which plays the same role as before; @continue@, which is called from
@drive_step@ when encountering an atomic expression, and handles the reduction context
(@Frame@) of the analyzed language; and @ndcontinue@, which dispatches meta-continuations
@NDCont@, and which is invoked from @continue@ when the normal stack is empty.

@@

Refunctionalization transforms first-order data types representing evaluation contexts 
to higher-order functions, and associates the dispatching logic of continuations with
proper higher-order functions.
This transformation can be also regarded as making the control flows of the defined 
language be the part of control flows of the meta-language.
%allocating continuation functions of the defined language in the heap of the meta-language.

@@

For clarity, we first show how to refunctionalize continuations and
meta-continuation without collecting semantics.
Then, adding an appropriate caching algorithm to make sure the analysis will
terminate at some fixed-point is a simple transformation of the program to cache-passing
style. Finally, we examine how computable pushdown control-flow analysis is established
through refunctionalization and caching.

@@

Given a collection of elements of type @T@, an initial accumlator of type 
@Ans@, a function @k@ that works on element's type @T@ and takes a continuation 
of type @Ans=>Ans@, and a function @m@ that works on and returns a value of type 
@Ans@ (which is also type @Ans=>Ans@), @nd@ chooses an element in the collection 
and applies @k@ to it, accumulates the value (@acc++ans@ where the @ans@ is the 
result comes from @k@) and eventually applies function @m@ on the accumulated 
value when there is no element in the collection.

@@

\textit{Cache.}
Recall that the latent assumption of the worklist algorithm in classical AAM is that 
if we have seen a state @s@, it means we also have seen all the successors of state @s@.
The caching algorithm we adopt here replays the same assumption but in a big-step 
manner: if we have seen the configuration @c@, then it means we also have seen the 
values that are evaluated from the configuration @c@.
Here we will apply the fixed-point caching algorithm as described from
\citeauthor{darais2017abstracting}'s ADI paper \cite{darais2017abstracting}.

@@

Upon entering @aeval@, we first determine whether the @out@ cache
contains some values for the current configuration and use them immediately (if present).
This corresponds to small-step AAM with a worklist: when we have seen a state,
we can simply discard the state and continue working through the rest of the worklist @todo@.

@@

\textit{The Problem with Return-flows.}
Pushdown control-flow is a property in analysis that precisely models the runtime
call structure of the analyzed program. A pushdown control-flow analysis provides an 
as-exact-as-runtime return-flow when analyzing a program, but traditional 
control-flow analysis collapses the state space into a finite space and thus 
causes imprecise stack modeling.

@@

We have multiple choices of how to proceed:
\begin{itemize}
  \item By monadifying the continuations, we obtain the abstract definitional interpreters 
    in monadic style which are what \citeauthor{darais2017abstracting} describe \cite{darais2017abstracting}.
  \item By representing the extended continuation-passing style with delimited control operators
    such as @shift@ and @reset@, we derive a new form of abstract interpreters in direct-style.
  \item In fact, we may also just use the same caching algorithm but with side effects such as
    assignments and mutations to update the cache, and then achieve the same definitional 
    interpreter with pushdown control-flows. \citeauthor{friedman-ai} in a tutorial on abstract 
    interpreters use this style to model caches and updates \cite{friedman-ai}.
    In this case, nondeterminism can be easily handled via @for@ comprehension. 
\end{itemize}

@@

\textit{Abstract Interpretation and Control-Flow Analysis.}
\citeauthor{cousot1977abstract} invented abstract interpretation as a sound approach to
approximate a program's runtime behavior \cite{cousot1977abstract}. Control-flow analysis
is one instance of abstract interpretation on functional programs that can be traced to
Jones \cite{jones1981flow}. \citeauthor{Shivers:1988:CFA:53990.54007} introduced $k$-CFA
which uses $k$ recent calling contexts as program contours that differentiate values from
different contexts \cite{Shivers:1988:CFA:53990.54007, Shivers:1991:SSC:115865.115884}.
The recent formulation of control-flow analysis is the abstracting abstract machines
methodology \cite{van2012systematic, van2010abstracting} which forms the starting point
of this paper. Modern programming languages such as Java \cite{might2010resolving} and
Racket \cite{Tobin-Hochstadt:2012:HSE:2384616.2384655} can also be modeled by the abstracting
abstract machine approach. \\

\textit{Defunctionalization and Refunctionalization.}
Defunctionalization and refunctionalization build connections between abstract machines and
interpreters. \citeauthor{Reynolds:72} first shows defunctionalization as a program transformation 
technique that can be used to transform higher-order functions to first-order functions when
writing a definitional interpreter \cite{Reynolds:72, Reynolds:HOSC98-revisited}.
Refunctionalization~\cite{TODOrefunc} is the left inverse of this transformation.

Over the years, Danvy and his collaborators applied refunctionalization and defunctionalization 
to many different abstract machines and interpreters and build functional correspondences among
them \cite{Ager:2003:FCE:888251.888254, Danvy:2001:DW:773184.773202, danvy2004refocusing,
Danvy:2008:DIP:1411204.1411206, AGER2004223, ager2005functional, Danvy:2006:RW:2171265.2171268,
DANVY2009534, danvy2009towards, biernacka2009towards}. Such semantic artifacts include the CEK machines,
the CLS machines, the SECD machines, call-by-need evaluators, monadic evaluators, etc.
Other applications are also found to be applicable, such as Dyck word recognizer, and 
Dijkstra's shunting-yard algorithm and even Quicksort algorithm \cite{Danvy:2006:RW:2171265.2171268,
DANVY2009534}.
For deterministic languages, \citeauthor{Danvy:2008:DIP:1411204.1411206} shows a reduction semantics
is a structural operational semantics in continuation-passing style, where the reduction contexts
are defunctionalized (data type representing) continuations \cite{Danvy:2008:DIP:1411204.1411206}.
Not unexpectedly, the correspondence also exists for nondeterministic languages. For example,
\citeauthor{Danvy:2001:DW:773184.773202} show that refunctionalizing a regular expressions matcher yields
its counterpart with two layer of continuations \cite{Danvy:2001:DW:773184.773202}.
And the multi-layer continuation-passing style is the underlying semantics of control operators
like @reset@ and @shift@ \cite{Danvy:1990:AC:91556.91622}.

Our work applies these ideas to the nondeterministic AAM. The additional layer of continuation controls
the nondeterministic choice over closures. After identifying and explicitly exposing (Section~\ref{linear})
that layer of meta-continuations, the whole program can be refunctionalized to extended
continuation-passing style, and then back to direct-style by the use of delimited control operators.

Lightweight fusion as one of the transformation in our paper is also studied by 
\citeauthor{Ohori:2007:LFF:1190216.1190241} \cite{Ohori:2007:LFF:1190216.1190241}.
As shown by \citeauthor{DANVY2008100}, lightweight fusion can be used to show the 
equivalence between small-step and big-step abstract machines \cite{DANVY2008100}.\\

>>>>

%% Title information
\title{Refunctionalization of Abstract Abstract Machines (Functional Pearl)}         %% [Short Title] is optional;
                                        %% when present, will be used in
                                        %% header instead of Full Title.
%\titlenote{with title note}             %% \titlenote is optional;
                                        %% can be repeated if necessary;
                                        %% contents suppressed with 'anonymous'
\subtitle{Bridging the Gap Between Abstract Abstract Machines and Abstract Definitional Interpreters}        %% \subtitle is optional
%\subtitlenote{with subtitle note}       %% \subtitlenote is optional;
                                        %% can be repeated if necessary;
                                        %% contents suppressed with 'anonymous'

@@

  In this paper, we show their functional correspondence and how to syntactically 
  transform small-step abstract abstract machines into big-step abstract definitional
  interpreters.
  The transformations include linearization, lightweight fusion, disentanglement, 
  refunctionalization, and finally from CPS back to direct-style with delimited 
  control operators. 
  Linearization expresses nondeterministic choices by first-order data types, 
  after which refunctionalization transforms the first-order data types representing continuations
  into higher-order functions. 
  The refunctionalized AAM is an abstract interpreter written in continuation-passing style
  with two layers of continuations.

@@

Applying the same idea to monadic definitional interpreters, \citeauthor{darais2017abstracting}
built monadic abstract definitional interpreters (ADI) in big-step style
\cite{darais2017abstracting}. One of the advantages of a monadic interpreter is
that it is modular and composable. By changing the underlying monads, the definition
of the interpreter is not modified, but we can recover different semantics, including
the concrete semantics and various abstract semantics such as context-sensitivity and
abstract garbage collection \cite{Sergey:2013:MAI:2491956.2491979}.

@@

In the domain of abstract semantic artifacts, by contrast, the relations between small-step
abstract abstract machines and big-step abstract definitional interpreters, as well as the
question of deriving one from the other, are not well studied.
One of the fundamental differences between concrete semantic artifacts and abstract semantic
artifacts is that a sound instance of the latter must tell you facts about all possible execution 
paths, and thus introduce nondeterminism. In addition to ensuring termination, abstract semantic
artifacts are usually equipped with a cache of reachable states; the cache is
iteratively updated in a monotonic way and guaranteed to reach the least fixed-point eventually.
In this functional pearl, we make a first step in answering the question of interderiving
abstract semantics artifacts, and provide a constructive answer to relate AAM and ADI
by presenting a series of syntactic transformations on the program from small-step
abstract abstract machines to big-step abstract definitional interpreters.

@@

  \item Section~\ref{disen} discusses \textbf{disentanglement}. In this part, we identify 
    different first-order data types which represent different layers of continuations, and
    disassemble their handlers into separate functions.

  \item Applying \textbf{refunctionalization} \cite{DANVY2009534, Danvy:2006:RW:2171265.2171268} 
    to disentangled AAM is shown in Section~\ref{sec:refunc}. 
    %as the left inverse of Reynolds' defunctionalization \cite{Reynold:72,Reynolds:HOSC98-revisited} 
    Refunctionalization replaces the first-order data types representing continuations with
    higher-order functions, and associates dispatching logic with proper higher-order functions. 
    We then derive an abstract interpreter written in continuation-passing style with two
    layers of continuations due to the nondeterminism.

@@

We can now faithfully describe the state transition function @step@,
which when given a machine state, determines its successor state.
The function @step@ is a partial function that only handles non-final states,
which must have a successor; the final case of a state is handled by function
@drive@, which will be mentioned in the next section.

@@

\subsection{Abstracting Abstract Machines} \label{aam}
Abstracting abstract machines (AAM) is a systematic methodology that derives sound
abstract interpreters for higher-order functional languages from concrete abstract 
machines \cite{van2012systematic, van2010abstracting}. An abstract abstract machine
implements computable abstract semantics which approximates the run-time behaviors of 
programs.
Since the state space of concrete execution is possibly infinite, the key insight of 
AAM approach when analyzing programs is to allocate both bindings and continuations 
on the store, and then bound the addresses space to be finite. Since each component
of state is finite, the abstracted machine-state space is also finite, and therefore
computable.

@@

For clarity, we divide the store into two separate stores: the binding stores @BStore@, and
the continuation store @KStore@.
The binding store maps binding addresses to sets of closure values, whereas the continuation
store maps continuation addresses to sets of continuations.
We then define a generic class @Store[K,V]@ that performs joining when updating elements
in a store (below). By parameterizing @Store[K,V]@ with @[BAddr, Storable]@ and
@[KAddr, Cont]@, we obtain @BStore@ and @KStore@, respectively.
We note that both the value store and continuation store are updated monotonically;
it continuously grows and never shrinks.

@@

The co-domain of binding stores @Storable@ is the same as previously defined for CESK 
machines. The conman of continuation stores @Cont@, on the other hand, is comprised of
a @Frame@ object and a continuation address @KAddr@. To mimic the run-time call stack, 
@KAddr@ plays the role of representing the remaining stack frames.
But since the continuation store may contain multiple continuations, the dereferencing 
of continuation addresses is also nondeterministic.

@@

\subsubsection{Allocating Addresses}
Up to this point, we have not described allocating addresses in stores, nor handling
the time stamp @Time@. In abstract interpretation, however, these are key ingredients
to achieve analyses with different sensitivities, as well as to perform a finite
state space analysis \cite{Gilray:2016:ACP:2951913.2951936}.
To effectively approximate the run-time behavior, we introduce a finite program contours
@time@ that encodes the program execution history. The function @tick@ is used to refresh
the "time" and get a "new time". We use a list of expression (which are drawn from the
control component) to encode the calling contexts history, and as we will see in Section~
\ref{kcfainst}, by applying different @tick@ functions on the timestamp, we are able to
obtain a family of analyses.

@@

However, in contrast to the concrete CESK machine, the @drive@ function performs 
collecting semantics instead of the valuation semantics. That is, for the purpose of
analyzing programs, the function @drive@ collects all the intermediate machine states 
as the program is abstractly executing. The following code shows a variant of the 
work-list algorithm to find the fixed-point of the set of states.
Function @drive@ always applies function @step@ to the head element @hd@ of
the work-list @todo@ if @hd@ is unseen. It then inserts the result of @step@ to
the rest of work-list, and in the meantime adds @hd@ to the explored states set.
If the work-list is empty, @drive@ simply returns the set of reachable states
up to the current execution point.

@@

In this section, we describe a variant of AAM that allows the stack to be unbounded 
which uses a precise model of call stack as we did in the concrete CESK machine.
Instead of allocating continuations in the store and embedding addresses of 
continuations in states, we intend to use a list of frames to explicitly model 
the stack. By doing so, we recover the call structure as the same as run-times
(so called pushdown control-flow analysis), but since the stack is unbounded, 
the state space is potentially infinite.

The pushdown AAM with an unbounded stack is also described by \citeauthor{van2012systematic} 
\cite{van2012systematic}. 
The reason we show it here is to keep the consistency of abstract semantics during 
transformations -- since the eventual abstract definitional interpreters simply inherit
a precise control-flow from its defining language, we also would like to start from a
pushdown AAM that is also able to precisely match calls and returns.
For readers who are not familiar with pushdown analysis, we have a detailed discussion 
in Section~\ref{pdcfarevisit}.

@@

In the previous section, we show that by keeping an unabstracted stack in the state
space, we can recover the precise call/return match. Now we begin describing the
transformations step by step. Our base machine is the AAM with an unbounded
stack, as we will later transform the stack structure to higher-order functions
representing continuations, it does not matter what kind of AAM we start from.
This is because it would not be possible to construct a new frame on the stack, nor
to allocate continuations in the store after Refunctionalization.
Thus, our choice to start from an unbounded-stack AAM is motivated simply because
it has an equivalent stack model to abstract definitional interpreters (our final target).

The underlying semantics of AAM is fundamentally nondeterministic: there are possibly
multiple target closures when dereferencing an address, and we have to explore both branches
if conditionals exist in the analyzed language. But the reduction context (i.e., the @Frame@ in
our program) is \emph{not} nondeterministic, and actually the work-list @todo@ implicitly
handles the nondeterminism. Thus, if we simply refunctionalize the frames to higher-order
functions, it does not help us to move toward abstract definitional interpreters.

@@

We apply this idea to AAM and our first step of transformations is to linearize all
nondeterministic choices. This step removes all nondeterminism from the @step@ function.
Likewise, the @Frame@ saves the information of the caller in concrete executions.
We then introduce another layer of control, and define a case class @NDCont@
that saves the information at a fork point when we have nondeterministic target closures.
We also add a new field @ndk@ of type @NDCont@ to the definition of state, which we now 
call @NDState@. For clarity of presentation in differentiating between the two continuations, 
we elect to call the first a \emph{continuation}, and the second a \emph{meta-continuation}
\footnote{Traditionally, they are also called \emph{success continuation} and \emph{failure
continuation} \cite{Danvy:1990:AC:91556.91622}.}.

@@

Here we also use first-order data types to explicitly represent the meta-continuation which
controls the nondeterminism.
Each @NDCont@ object contains a list of closures that are possible functions to be invoked,
a set of values that will be bound to the target function's formal argument, and the store, time,
and frames at the fork point. In the definition of @NDState@, an auxiliary function @toState@
is added that converts itself to type @State@.

@@

By atomically evaluating @f@, we obtain a set of closures, with the transition being 
deterministic in regards to the first element of the closure set.
As such, we prepare a new environment, a new store, and a new frame list only for the
first closure in that set. A new meta-continuation @new_ndk@ is also constructed,
which contains the rest of closures, the values of argument @ae@, the store, the time,
and the new frame list. We use the store before updating, because for different closures
they may form different binding addresses @baddr@. We also use the new frames, because
all closures at this fork point share the same stack structure and return point.

@@

Normally, if the continuation @konts@ is an empty list, then it means we have reached
the end of the computation (i.e., a halt); otherwise, we should return the values of 
@ae@ to its caller which is contained in the top frame of stack.

However, since we have added nondeterminism into state, an empty list of frames means
we have reached the halt of \textit{one computation path}, and we should determine the
next state indicated by @NDCont@. Thus, we have a inner pattern matching on @ndk@:

\begin{itemize}
  \item If the closure set is an empty set (@Nil@ in the code since we list to represent that set),
then we have tried all the closures of this fork point and should move to the
next fork point. Since the meta-continuation @nkd@ is represented by a list, and we always
append new elements to its front, we are resuming to the most recent fork point by
popping up the front-most element and using the tail of that list.
  \item Otherwise, we pop a closure from the set, then build a new
environment and a new store for the body expression of the closure;
we use the same frame list and time that are copied from the fork point rather
than current one.
The meta-continuation @new_ndk@ is also updated by the removal of that closure, which makes
sure we will eventually shrink to the empty set case.
\end{itemize}

@@

The @drive@ function is also changed: there is no work-list anymore, as all the
potentially unexplored states are embedded into the continuation for
nondeterminism.
The termination of the analysis occurs when we reach an @NDState@ in which the expression
is atomic and both the continuation and meta-continuation are empty lists. This corresponds 
to the case that @todo@ list is empty. The set of explored states is preserved as before.

At this point, we have obtained a linearized abstract abstract machine. If we imagine that the
classical AAM explores a graph of reachable states, then the linearized AAM flattens the
graph to a linear sequence of states.

@@

With fusing complete, our AAM appears similar to a ``big-step'' interpreter, although it
still has the machine state representation inside. In the disentanglement transformation,
we identify the first-order data types which represent reduction contexts
and lift the code blocks that handle these reduction contexts to be individual functions.
After disentanglement, we will obtain an abstract abstract machine that in defunctionalized form.

Since there are two layers of continuations, we disentangle them into three mutually recursive 
functions:
@drive_step@, which plays the same role as before; @continue@, which is called from
@drive_step@ when encountering an atomic expression, and handles the reduction context
(@Frame@) of the analyzed language; and @ndcontinue@, which dispatches meta-continuations
@NDCont@, and which is invoked from @continue@ when the normal stack is empty.

@@

Refunctionalization transforms first-order data types representing evaluation contexts 
to higher-order functions, and associates the dispatching logic of continuations with
proper higher-order functions.
After this transformation, we will obtain 
This transformation can be also regarded as making the control flows of the defined 
language be the part of control flows of the meta-language.
%allocating continuation functions of the defined language in the heap of the meta-language.

@@

For clarity, we first show how to refunctionalize continuations and meta-continuation 
without collecting semantics.
Then, adding an appropriate caching algorithm to ensure the analysis will
terminate at some fixed-point is a simple transformation of the program to cache-passing
style. Finally, we examine how computable pushdown control-flow analysis is established
through refunctionalization and caching.

@@

Given a collection of elements of type @T@, an initial accumulator of type 
@Ans@, a function @k@ that works on element's type @T@ and takes a continuation 
of type @Ans=>Ans@, and a function @m@ that works on and returns a value of type 
@Ans@ (which is also type @Ans=>Ans@), @nd@ chooses an element in the collection 
and applies @k@ to it, accumulates the value (@acc++ans@ where the @ans@ is the 
result comes from @k@) and eventually applies function @m@ on the accumulated 
value when there is no element in the collection.

@@

\textit{Cache.}
Recall that the latent assumption of the work-list algorithm in classical AAM is that 
if we have seen a state @s@, it means we also have seen all the successors of state @s@.
The caching algorithm we adopt here replays the same assumption but in a big-step 
manner: if we have seen the configuration @c@, then it means we also have seen the 
values that are evaluated from the configuration @c@.
Here we will apply the fixed-point caching algorithm as described from
\citeauthor{darais2017abstracting}'s ADI paper \cite{darais2017abstracting}.

@@

Upon entering @aeval@, we first determine whether the @out@ cache
contains some values for the current configuration and use them immediately (if present).
This corresponds to small-step AAM with a work-list: when we have seen a state,
we can simply discard the state and continue working through the rest of the work-list @todo@.

@@

\textit{The Problem with Return-flows.}
Pushdown control-flow is a property in analysis that precisely models the run-time
call structure of the analyzed program. A pushdown control-flow analysis provides an 
as-exact-as-run-time return-flow when analyzing a program, but traditional 
control-flow analysis collapses the state space into a finite space and thus 
causes imprecise stack modeling.

@@

We have multiple choices of how to proceed:
\begin{itemize}
  \item By representing the continuations with monads\cite{filinski1994representing}, 
    we obtain the abstract definitional interpreters in monadic style which are similar to what 
    \citeauthor{darais2017abstracting} describe \cite{darais2017abstracting}.
  \item By representing the extended continuation-passing style with delimited control operators
    such as @shift@ and @reset@, we derive a new form of abstract interpreters in direct-style.
  \item In fact, we may also just use the same caching algorithm but with side effects such as
    assignments and mutations to update the cache, and then achieve the same definitional 
    interpreter with pushdown control-flows. \citeauthor{friedman-ai} in a tutorial on abstract 
    interpreters use this style to model caches and updates \cite{friedman-ai}.
    In this case, nondeterminism can be easily handled via @for@ comprehension. 
\end{itemize}

@@

\textit{Defunctionalization and Refunctionalization.}
\citeauthor{Reynolds:72} first shows defunctionalization as a program transformation 
technique that can be used to transform higher-order functions to first-order data types and 
its first-order dispatching function when writing definitional interpreters
\cite{Reynolds:72, Reynolds:HOSC98-revisited}. As generalized closure conversion,
defunctionalization is also used in program analysis, transformation, and partial evaluation. %TODO cite
Refunctionalization~\cite{Danvy:2006:RW:2171265.2171268, Danvy:2001:DW:773184.773202} is the 
left inverse of defunctionalization, firstly considered by \citeauthor{Danvy:2001:DW:773184.773202}.

\citeauthor{Ager:2003:FCE:888251.888254} revealed that defunctionalization and refunctionalization
can be used to show the functional correspondence between interpreters and abstract machines
\cite{Ager:2003:FCE:888251.888254}. A number of independently designed small-step semantic artifacts can 
be obtained by closure-converting, CPS-transforming, and then defunctionalizing their big-step counterparts, 
and vice versa by refunctionalizing \cite{DANVY2009534}. 
Such semantic artifacts include SECD machines \cite{10.1007/11431664_4, Ager:2003:FCE:888251.888254, DanvySECDJ},
CEK machines \cite{Ager:2003:FCE:888251.888254}, CLS machines  \cite{Ager:2003:FCE:888251.888254},
Categorical Abstract Machines \cite{Ager:2003:FCE:888251.888254}, lazy abstract machines \cite{AGER2004223}, 
monadic evaluators \cite{ager2005functional}, as well as languages with richer constructs
\cite{Danvy:2008:DIP:1411204.1411206, danvy2009towards, biernacka2009towards}.

Besides, Danvy and his collaborators observed the correspondence between evaluation contexts and continuations.
For deterministic languages, \citeauthor{Danvy:2008:DIP:1411204.1411206} showed that reduction contexts
are defunctionalized (data type representing) continuations and reduction semantics is a structural 
operational semantics in continuation-passing style \cite{Danvy:2008:DIP:1411204.1411206}.
Not unexpectedly, the correspondence also exists for nondeterministic languages. For example,
\citeauthor{Danvy:2001:DW:773184.773202} showed that refunctionalizing a regular expression matcher yields
its counterpart with two layer of continuations \cite{Danvy:2001:DW:773184.773202}.
And the operational semantics of control operators (e.g., @reset@ and @shift@) can be given by an 
abstract machine with multi-layer continuations \cite{Danvy:1990:AC:91556.91622, lmcs:2269}.

More generally, other kinds of programs, instead of merely semantic artifacts, are also found to be applicable.
As \citeauthor{Danvy:2001:DW:773184.773202} pointed out, defunctionalization is dual to Church encoding.
Examples include backtracking algorithms (e.g., regular expression matcher), Dyck word recognizer, and 
Dijkstra's shunting-yard algorithm and even Quicksort algorithm \cite{Danvy:2006:RW:2171265.2171268, 
DANVY2009534, Danvy:2001:DW:773184.773202}. \\

%Our work applies these ideas to the nondeterministic AAM. The additional layer of continuation controls
%the nondeterministic choice over closures. After identifying and explicitly exposing (Section~\ref{linear})
%that layer of meta-continuations, the whole program can be refunctionalized to extended
%continuation-passing style, and then back to direct-style by the use of delimited control operators. \\
